Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
reorg
subtype

\paragraph{Nullables} 
The type system has a \null type inhabited by a singleton \code{NULL}
value. This value is often used as a sentinel or as a stand in for arguments
that were not provided.  To capture this behavior, we introduce a nullable
type \code{?}\T.  The difference between the \NA and \code{NULL} values is
that \NA can be stored in primitive vectors.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related. Any vector can be converted to a list
with \code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). We support parametrization of the list type, \LIST\T, but for
heterogeneous lists this type will default to \any.

@@

\paragraph{Environments}
Environments are lists with reference semantics: mutating a value in an
environment is performed in-place.  They are used to store variables and to
escape from the copy-on-write semantics of other data types.  Environments
are comprised of name, value pairs, and each environment has a reference to
its parent environment.

%
%
\section{Extracting and Checking Signatures}

@@

\medskip
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
    $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\  
\end{tabular}
\medskip

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types. Lists with more than 5 types in their union become \LIST{any}.
We also only generate types for the first 20 arguments of a function.

@@

@@

\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}~\VEC\raw

\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}? ~\VEC\chr

\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}~\VEC\raw

\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}~\VEC\dbl

>>>>

\paragraph{Nullables} 
The \null type is inhabited by a singleton \code{NULL} value often used as a
sentinel or a stand in for arguments with no value.  To capture this
behavior, we introduce a nullable type \code{?}\T.  The difference between
the \NA and \code{NULL} values is that the latter cannot be stored in
vectors.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related: a vector converts to a list with
\code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). The list type is parameterize, \LIST\T, when the list is
heterogeneous \T defaults to \any.

@@

\paragraph{Environments}
Environments are lists with reference semantics: mutating a value in an
environment is performed in-place.  They are used to store variables and to
escape from the copy-on-write semantics of other data types.

\section{Extracting and Checking Signatures}

@@

\medskip
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
  \\
  \\
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\medskip

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types. For performance reasons, we bound lists to no
more than five types in their union, above that they become \LIST{any}.
Functions are alos limited to twenty arguments.

@@

\code{openssl::decrypt\_envelope}: \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw, \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw}

\code{dbplyr::set\_win\_current\_group} \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr}

\code{openssl::sha256}: \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw}

\code{forecast::initparam}: \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl, ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl}

@@
